Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-30 submission/manuscript.Rmd
update abstract

**Background.** Colorectal cancer (CRC) is a growing health concern with the majority of the risk for developing disease being due to environmental factors. The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC, while other taxa are associated to colons without tumors. Some of the taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate. Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). There was no difference in classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).

**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC. Alternatively, there may be other mechanisms that have not been thoroughly investigated that are more important to the development of human CRC.

>>>>

**Background.** Colorectal cancer (CRC) is a continuing health concern with the majority of the risk for developing disease being due to environmental factors. The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC and other taxa being associated to colons without tumors. Some of the bacterial species in taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate. Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70). Finally, there was no difference between classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).

**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC. Alternatively, our observations also support the hypothesis that there may be other mechanisms that have not been thoroughly investigated that are more involved with the development of human CRC.

### 2018-07-30 submission/manuscript.tex
update abstract

\textbf{Background.} Colorectal cancer (CRC) is a growing health concern
with the majority of the risk for developing disease being due to
environmental factors. The microbiota is one of these environmental
factors with certain bacterial community members being associated with
CRC, while other taxa are associated to colons without tumors. Some of
the taxa associated to colons without tumors can use fiber to produce
short-chain fatty acids (SCFAs) that can inhibit tumor growth in model
systems. However, the data supporting the importance of SCFAs in human
CRC is less certain. Here, we test the hypothesis that SCFA
concentrations are different in individuals with colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their concentrations of
acetate, butyrate, and propionate. Analysis also included tumor
classification models using Random Forest, imputed gene relative
abundance with PICRUSt, and metagenomic sequencing on a subset (n=85) of
the total cross-sectional group.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). There was no difference in
classification models with or without SCFAs in their ability to predict
patients with adenomas or carcinomas versus individuals without tumors
(P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
also no difference in genes involved with SCFA synthesis between
individuals without tumors and patients with adenomas or carcinomas
(P-value \textgreater{} 0.70).

\textbf{Conclusions.} Although our data does not support the hypothesis
that SCFAs are different in individuals that have colorectal tumors,
there may be context specific scenarios where SCFAs may still be
beneficial for treatment of CRC. Alternatively, there may be other
mechanisms that have not been thoroughly investigated that are more
important to the development of human CRC.

>>>>

\textbf{Background.} Colorectal cancer (CRC) is a continuing health
concern with the majority of the risk for developing disease being due
to environmental factors. The microbiota is one of these environmental
factors with certain bacterial community members being associated with
CRC and other taxa being associated to colons without tumors. Some of
the bacterial species in taxa associated to colons without tumors can
use fiber to produce short-chain fatty acids (SCFAs) that can inhibit
tumor growth in model systems. However, the data supporting the
importance of SCFAs in human CRC is less certain. Here, we test the
hypothesis that SCFA concentrations are different in individuals with
colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their concentrations of
acetate, butyrate, and propionate. Analysis also included imputed gene
relative abundance with PICRUSt, metagenomic sequencing on a subset
(n=85) of the total cross-sectional group, and tumor classification and
SCFA prediction models using Random Forest.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
there was also no difference in genes involved with SCFA synthesis
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.70). Finally, there was no
difference between classification models with or without SCFAs in their
ability to predict patients with adenomas or carcinomas versus
individuals without tumors (P-value \textgreater{} 0.05).

\textbf{Conclusions.} Although our data does not support the hypothesis
that SCFAs are different in individuals that have colorectal tumors,
there may be context specific scenarios where SCFAs may still be
beneficial for treatment of CRC. Alternatively, our observations also
support the hypothesis that there may be other mechanisms that have not
been thoroughly investigated that are more involved with the development
of human CRC.
